---
title: "Stylistic Variations in Neil Gaiman's Work: A Corpus-Driven Analysis"
author: "Hongxu Zhou" 
format: 
    pdf:
        toc: true
        number-sections: true
        reference-location: document
        mainfont: "Times New Roman"
        fontsize: 12pt
        linestretch: 1.5
bibliography: corpus_ling_ref.bib
csl: apa.csl
---

# Abstract

# Introduction

As a prominent contemporary author, Neil Gaiman has produced a substantial body of work spanning diverse genres and targeting audiences ranging from children to adults. This project employs corpus-driven methods to analyse potential stylistic variations in Gaiman's writing that may distinguish works intended for different reader demographics. Through systematic linguistic analysis, the study aims to identify and characterize textual features that differentiate Gaiman's writing across audience-targeted works.

```{r}
#| label: packages_loading
#| message: false
#| warning: false
#| include: false
#| paged-print: false
# Load required packages
library(tidyverse)      
library(here)           
library(quanteda)       
library(quanteda.textstats)  
library(sentimentr)     
library(wordcloud)      
library(ggplot2)        
library(gridExtra)     
library(scales)        
library(ggridges)      
library(dplyr)
library(stopwords)
library(udpipe)
library(ggrepel) 
library(factoextra) 
```

```{r}
#| label: corpus_loading
#| include: false

# Read the corpus
gaiman_corpus <- readRDS(here("corpora", "gaiman_corpus_complete.rds"))

# Explore the structure
str(gaiman_corpus, max.level = 1)
```

The corpus contains three levels of the texts: document-level (tokenised by chapter); sentence-level (tokenised by sentence-end punctuation); and token-level (tokenised by word).

\
To accelerate the calculation and analysis, some basic statistical features are calculated beforehand and added to the corpus as the layers of Metadata and Vocabulary.

# Literature Review

-   Previous corpus-based studies on stylistic variation in fiction

-   Research on linguistic markers of audience targeting (children's vs. adult literature)

-   Studies on Neil Gaiman's writing style (if available)

-   Research aim: to find out if it is possible to differentiate the targeted audience demographics through a pure data-driven approach.

-   **Drafted Research Questions**:

    1.  What linguistic features differentiate Gaiman's writing across audience demographics?

# Methodology

## Corpus

This study employs a specialised corpus comprising ten books by Neil Gaiman, selected to represent the author's rage across audience demographics. The ten books are:

| Book Title                       | Abbr. | Goodreads Genre Tags    |
|----------------------------------|-------|-------------------------|
| American Gods                    | AG    | Adult                   |
| Anansi Boys                      | AB    | Adult                   |
| Neverwhere                       | NW    | Adult; Young-Adult      |
| Smoke and Mirrors                | SM    | Adult; Young-Adult      |
| The Ocean at the End of the Lane | OCEAN | Young-Adult; Adult      |
| Coraline                         | CL    | Young-Adult; Children;  |
| The Graveyard Book               | TGB   | Young-Adult; Children   |
| Fortunately, the Milk            | FTM   | Children, Young-Adult   |
| Odd the the Frost Giants         | OFG   | Children, Young-Adult   |
| Stardust                         | ST    | Young Adult, Adult      |

: Books in Gaiman Corpus

They span Gaiman's career and represent various genres within his oeuvre. Therefore, this corpus is a balanced representation for analysis of potential stylistic variations.

Each books was obtained in EPUB format, which preserved textual content while maintaining structural elements that enables programming analysis. The `epubr` package [@leonawicz2025] was used to extract and parse content from these books while preserving metadata and chapter boundaries.

The the cop was designed with the multi-tired structure to accelerate analysis at different levels of linguistic granularity:

1.  **Document-level**: Each chapter/single story constitute a document. This layer of corpus keeps narrative unites as defined by the author, and so enables analysis of chapter-level stylistic patterns and narrative structure.
2.  **Sentence-level**: The text is segmented into individual sentences using the `tokenizers` package [@mullen], which applies rule-based sentence boundary disambiguation. This intermediate level allows for analysis of syntactic patterns and sentence complexity.
3.  **Token-level**: Individual words are extracted for lexical analysis. Tokens maintain references to their source sentences and documents, enabling multi-level analysis.

Additionally, two supplementary components are created:

1.  **Metadata layer**: It contains corpus-level statistics (total size, unique vocabulary) and book-level metadata (lexical diversity, average sentence length).
2.  **Vocabulary layer**: It provides frequency and distribution for each unique token in the corpus.

As a result, the corpus contains 711,385 words across 154 documents (chapter/story), with an average of 15.4 chapters per book. The corpus contains 60,127 sentences with an average length of 12.3 words per sentence. The vocabulary size (type) is 28, 968.

```{r}
#| label: exploratory_analysis
#| message: false
#| warning: false
#| include: false
# Extract book-level metadata
book_metadata <- gaiman_corpus$metadata$book_level

# Display basic statistics about the corpus 
corpus_summary <- book_metadata |> 
  summarise(
    total_books = n(),
    total_words = sum(total_words),
    avg_words_per_book = mean(total_words),
    avg_chapters = mean(chapters),
    avg_sentences = mean(sentences),
    avg_sentence_length = mean(avg_sentence_length),
    avg_lexical_diversity = mean(lexical_diversity)
  )

print(corpus_summary)
```

## Analysis

### lexical level

The lexical analysis examines word-level patterns that might distinguish text for different audience groups. I analysed word length distribution, lexical diversity, and distinctive vocabulary to investigate potential stylistic variations.

#### Word Length Distribution

Word length often correlates with text complexity — longer words typically appear in more advanced or specialised text [@biber1988]. @fig-word_length_vis presents the percentage of words at each length across Gaiman's works. This visualisation reveals a remarkably consistent pattern across all books, with most words averaging around 4 characters regardless of the intended audience. This suggests that at the the basic word length level, Gaiman maintains a relatively stable style.

```{r}
#| label: word_length_cal
#| message: false
#| warning: false
#| include: false
#| paged-print: false
# Calculate word length for all tokens
token_length <- gaiman_corpus$token_level |> 
  # Calculate character length of each token 
  mutate(word_length = nchar(token))

# With token length, we can calculate word lengh by book
word_lengh_stats <- token_length |> 
  group_by(book_id) |> 
  summarise(
    avg_length = mean(word_length),
    median_length = median(word_length),
    sd_length = sd(word_length),
    # We can further check the percentage of words by length 
    short_word_pct = mean(word_length <= 4) * 100, # 1-4 words = short
    medium_word_pct = mean(word_length > 4 & word_length <= 8)* 100, # 4 - 8 = medium
    long_word_pct = mean(word_length >8)*100, # 9+ words are long
    .groups = "drop"
  ) |> 
  arrange(desc(avg_length))

print(word_lengh_stats)
```

```{r}
#| label: fig-word_length_vis
#| echo: false
#| fig-cap: Word Length Distribution by Book
word_lengh_dist <- token_length |> 
  # Count occurrences of each word length in each book 
  count(book_id, word_length) |> 
  # Group by book to calculate percentages 
  group_by(book_id) |> 
  mutate(percentage = n / sum(n) * 100) |> 
  ungroup() |> 
  # start vis
  ggplot(aes(x = word_length, 
             y = percentage,
             color = book_id,
             group = book_id)) +
  geom_line(linewidth = 1, alpha = 0.7) + # no more "size", update to linewidth
  labs(
    title = "Word Length Distribution by Book", 
    subtitle = "Percentage of words at each character length",
    x = "Word Length (by character)",
    Y = "Percentage of books",
    color = "Book"
    
  )+
  theme_classic() +
  theme(legend.position = "bottom")

print(word_lengh_dist)
```

#### Lexical Diversity

Lexical diversity is measured using the type-token ratio (TTR), which represents vocabulary richness:

$$
Lexical\quad Diversity = \frac {Number \quad of \quad Unique \quad Words} {Total \quad Word \quad Count}
$$

```{r}
#| label: fig-ttr_visualisation
#| echo: false
#| fig-cap: Lexical Diversity by Book
# Lexical diversity is already stored in the corpus, so we just need to visulaise it 
lex_div_plot <- ggplot(book_metadata, 
                      aes(x = reorder(book_id, lexical_diversity), 
                          y = lexical_diversity)) +
  geom_bar(stat = "identity", fill = "lightgreen", alpha = 0.8) +
  # Add reference line for average
  geom_hline(yintercept = mean(book_metadata$lexical_diversity), 
             linetype = "dashed", color = "red") +
  # Add text annotation for average
  annotate("text", 
           x = 1, 
           y = mean(book_metadata$lexical_diversity) + 0.01, 
           label = "Corpus Average", 
           hjust = 0, 
           color = "red") +
  labs(
    #title = "Lexical Diversity by Book",
    subtitle = "Higher values indicate more varied vocabulary",
    x = "Book",
    y = "Type-Token Ratio"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(lex_div_plot)

```

@fig-ttr_visualisation displays the lexical diversity values cross the 10 books. Interestingly, the TTR values do not clearly separate along audience lines. Even, Fortunately, the Milk, a typical picture book for children, has the highest TTR value. This finding may virtually challenge the assumption that books for younger readers use simpler vocabulary. However, it is acknowledged that TTR is sensitive to text length – longer texts naturally tend toward lower TTR values due to necessary repetition of function words and proper names. This limitation suggests the need for additional, more sophisticated linguistic measures beyond basic lexical metrics.

```{r}
#| label: word_frequnecy
#| include: false
vocabulary <- gaiman_corpus$vocabulary
stopwords = stopwords(language = "en")

top_content_words <- vocabulary |> 
  filter(!(token_lower %in% stopwords)) |> 
  arrange(desc(frequency)) |> 
  head(30)

# Skip printing resutls, directly visualise the bar chart
top_words_plot <- ggplot(top_content_words,
                         aes(x = reorder(token_lower, frequency),
                             y = frequency)) +
  geom_bar(stat = "identity", fill = "cornflowerblue") +
  labs(
    title = "Top 30 Conent Words in Gaiman Corpus",
    subtitle = "Stopwords applied",
    x = "Word",
    y = "Frequency"
  ) + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

print(top_words_plot)
```

```{r}
#| label: quanteda_loading
#| include: false
# Create a corpus from the document level data
quanteda_corpus <- corpus(
  gaiman_corpus$document_level$text,
  docnames = gaiman_corpus$document_level$doc_id
)

# Add document variables for grouping
docvars(quanteda_corpus, "book_id") <- gaiman_corpus$document_level$book_id
docvars(quanteda_corpus, "title") <- gaiman_corpus$document_level$title
docvars(quanteda_corpus, "chapter_num") <- gaiman_corpus$document_level$chapter_num

```

#### Keyword analysis

To identify vocabulary patterns distinctive to each book, I use a log-odds ratio analysis to compare word frequencies within each book against the rest of the corpus.

```{r}
#| label: log_off_cal
#| include: false
# Log odds function to compare one book against all others
# This identifies distinctive words in each book compared to other books
calculate_log_odds <- function(dfm_obj, target_book) {
  # Get index of target book
  target_idx <- which(docnames(dfm_obj) == target_book)
  
  # Skip if book not found
  if(length(target_idx) == 0) return(NULL)
  
  # Get counts for target book
  target_counts <- as.numeric(dfm_obj[target_idx,])
  target_total <- sum(target_counts)
  
  # Get counts for all other books
  others_counts <- colSums(dfm_obj[-target_idx,])
  others_total <- sum(others_counts)
  
  # Calculate log odds (adding small constant to prevent division by zero)
  epsilon <- 0.5  # Smoothing constant
  log_odds <- log((target_counts + epsilon) / (target_total - target_counts + epsilon)) - 
              log((others_counts + epsilon) / (others_total - others_counts + epsilon))
  
  # Create results data frame
  data.frame(
    term = colnames(dfm_obj),
    log_odds = log_odds,
    target_count = target_counts,
    others_count = others_counts,
    stringsAsFactors = FALSE
  ) %>%
    # Sort by log odds (descending)
    arrange(desc(log_odds))
}

# Create document-feature matrix
book_dfm <- tokens(quanteda_corpus) %>%
  tokens_remove(stopwords("en")) %>%
  tokens_remove(pattern = "[\\d\\p{P}]", valuetype = "regex") %>% # Remove numbers and punctuation
  dfm() %>%
  dfm_trim(min_termfreq = 3) %>%  # Remove very rare terms
  dfm_group(groups = docvars(quanteda_corpus, "book_id"))

# Get log odds for each book
book_ids <- unique(docvars(quanteda_corpus, "book_id"))
all_keywords <- list()

for (book_id in book_ids) {
  # Calculate log odds
  book_keywords <- calculate_log_odds(book_dfm, book_id)
  
  # Skip if book not found in DFM
  if(is.null(book_keywords)) next
  
  # Keep top 20 distinctive words
  book_keywords <- book_keywords %>%
    head(20) %>%
    mutate(book_id = book_id)
  
  # Add to results list
  all_keywords[[book_id]] <- book_keywords
}

# Combine results
keywords_df <- bind_rows(all_keywords)
```

```{r}
#| label: fig-log_off_keywords
#| echo: false
#| fig-cap: Top 10 Keywords by Book
# Visualize top 10 keywords for each book
top_keywords_plot <- keywords_df %>%
  group_by(book_id) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  ggplot(aes(x = tidytext::reorder_within(term, log_odds, book_id), 
             y = log_odds, 
             fill = book_id)) +
  geom_col() +
  facet_wrap(~ book_id, scales = "free_y") +
  upstartr::scale_x_reordered() +  # Required for reorder_within
  coord_flip() +
  labs(
    #title = "Top 10 Distinctive Words by Book",
    subtitle = "Based on log odds ratio compared to other books",
    x = "Term",
    y = "Log Odds Ratio"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(top_keywords_plot)
```

@fig-log_off_keywords shows the top distinctive words for each book. The keyword analysis reveals thematic differences more than stylistic ones: distinctive words primarily reflect characters and plot elements rather than systematic vocabulary differences based on audience age.

The lexical results suggest that Gaiman does not significantly simplify his vocabulary for younger readers. This finding may contraindicate conventional assumptions about children's literature. The unexpected pattern prompts to look beyond word-level features toward sentence structure and complexity, where more pronounced stylistic variations might emerge.

### Sentence Level

The sentence level analysis provides deeper insights into the stylistic variation in the corpus than word-level features alone. Based on sentence structure analysis, I examine complexity, readability and content characteristics across different works.

#### Sentence length analysis

```{r}
#| label: sent_loading
#| include: false

# Prepare sentence data
sentence_data <- gaiman_corpus$sentence_level %>%
  select(book_id, title, chapter_num, sentence_id, text)
```

```{r}
#| label: fig-sentent_length_dist_vis
#| echo: false
#| fig-cap: Sentence Length Distribution by Book
# Calculate sentence length statistics
sentence_length_stats <- sentence_data %>%
  # Add sentence length in words
  mutate(sentence_length = sapply(strsplit(text, "\\s+"), length)) %>%
  # Group by book
  group_by(book_id) %>%
  # Calculate statistics
  summarise(
    avg_length = mean(sentence_length),
    median_length = median(sentence_length),
    sd_length = sd(sentence_length),
    min_length = min(sentence_length),
    max_length = max(sentence_length),
    short_sent_pct = mean(sentence_length <= 5) * 100,  # Percentage of short sentences
    long_sent_pct = mean(sentence_length > 20) * 100    # Percentage of long sentences
  ) %>%
  arrange(desc(avg_length))

# Visualize sentence length distribution
sentence_length_vis <- sentence_data %>%
  # Add sentence length in words
  mutate(sentence_length = sapply(strsplit(text, "\\s+"), length)) %>%
  # Remove extreme outliers for better visualization
  filter(sentence_length <= 50) %>%
  ggplot(aes(x = sentence_length, y = book_id, fill = book_id)) +
  geom_density_ridges(alpha = 0.7, scale = 3) +
  labs(
    #title = "Sentence Length Distribution by Book",
    subtitle = "Ridgeline plot of word counts per sentence",
    x = "Words per Sentence",
    y = "Book"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(sentence_length_vis)
```

As a fundamental metric, sentence length offers initial insights into text complexity. @fig-sentent_length_dist_vis displays the distribution of sentence lengths across Gaiman's books. While all of them show similar central tendencies with peaks around 7-12 words per sentence, *Fortunately, the Milk* and *Odd and the Frost Giants* display narrower distributions with fewer long sentences. The more controlled sentence length variance suggests that they are more likely for children, though the differences are subtle and require deeper syntactic analysis.

#### Syntactic Complexity

Syntactic complexity is measured through dependency parsing, which captures the relationships between words in a sentence. Mean dependency length (MDL) is a method that calculates the average distance between syntactically related words. It represents language comprehension difficulty [@haitaoliu2008]. Higher MDL values indicate more complex sentence structures, for example, with embedded or non-adjacent relationships.

This analysis sampled 100 sentences from each book and parsed them using the Universal Dependency framework provided by `udpipe` [@straka2016]. @fig-ud_vis presents the mean dependency length for each work. Three books that are generally considered for adults, *American Gods, Neverwhere,* and *Anansi Boys* show consistently higher MDL values (from 3.5 to 4.2) compared to those typically for children and younger audiences (*Odd and the Frost Giants, Fortunately, the Milk*), which show lower values (from 2.4 to 3). This pattern suggests Gaiman systematically adjusts syntactic complexity based on intended audience.

```{r}
#| label: ud_cal
#| include: false


# 1. Function to calculate mean dependency length for parsed text
calculate_mean_dep_length <- function(parsed_data) {
  # For each sentence, calculate dependency lengths
  parsed_data %>%
    # Filter out root nodes and punctuation
    filter(head_token_id != 0, upos != "PUNCT") %>%
    # Calculate absolute distance between each word and its head
    mutate(
      # Convert IDs to numeric and calculate distance
      token_id_num = as.numeric(token_id),
      head_id_num = as.numeric(head_token_id),
      # Absolute difference is the dependency length
      dep_length = abs(token_id_num - head_id_num)
    ) %>%
    # Calculate mean dependency length
    summarize(
      mean_dep_length = mean(dep_length, na.rm = TRUE),
      median_dep_length = median(dep_length, na.rm = TRUE),
      max_dep_length = max(dep_length, na.rm = TRUE),
      deps_analyzed = n(),
      .groups = "drop"
    )
}

# 2. Process a sample of text from each book to calculate dependency lengths
# This avoids processing the entire corpus which could be time-consuming
set.seed(42) # For reproducibility in sampling
sample_size <- 100 # Sample 100 sentences per book for analysis

# Get sample sentences from each book - FIXED version
sampled_sentences <- sentence_data %>%
  # Split by book_id
  group_by(book_id) %>%
  # Use group_modify to properly access group size
  group_modify(~{
    # Get the minimum of sample_size or actual group size
    sample_size_to_use <- min(sample_size, nrow(.x))
    # Sample that many rows
    slice_sample(.x, n = sample_size_to_use)
  }) %>%
  ungroup()

# Create a unique identifier for each sampled sentence
sampled_sentences <- sampled_sentences %>%
  mutate(sample_id = paste0(book_id, "_", row_number()))

# Initialize dependency parsing model
# Note: This requires downloading a model - this block shows expected code flow
dep_model <- udpipe_download_model(language = "english-ewt", model_dir = getwd())
ud_model <- udpipe_load_model(dep_model$file_model)

# Process the sample sentences with udpipe for dependency parsing
# Using the sample_id as doc_id to maintain uniqueness
parsed_samples <- udpipe(
  x = sampled_sentences$text, 
  object = ud_model,
  doc_id = sampled_sentences$sample_id,  # Use our unique sample_id
  parallel.cores = 1
)

# Create a mapping between sample_id and book_id
sample_book_mapping <- sampled_sentences %>%
  select(sample_id, book_id, title) %>%
  distinct()

# 3. Calculate mean dependency length for each sentence, then aggregate by book
sentence_dependency_stats <- parsed_samples %>%
  # Group by doc_id (which is our sample_id)
  group_by(doc_id) %>%
  # Calculate dependency stats for each sentence
  group_modify(~calculate_mean_dep_length(.x))

# Join with our mapping to get book information
book_dependency_stats <- sentence_dependency_stats %>%
  # Join with mapping to get book_id
  left_join(sample_book_mapping, by = c("doc_id" = "sample_id")) %>%
  # Group by book to aggregate sentence-level stats
  group_by(book_id, title) %>%
  # Calculate book-level averages
  summarize(
    mean_dep_length = mean(mean_dep_length, na.rm = TRUE),
    median_dep_length = mean(median_dep_length, na.rm = TRUE),
    max_dep_length = max(max_dep_length, na.rm = TRUE),
    sentences_analyzed = n(),
    .groups = "drop"
  )

```

```{r}
#| label: fig-ud_vis
#| echo: false
#| warning: false
#| fig-cap: Syntactic Complexity in Gaiman's Works

# 4. Create visualization of mean dependency length
dep_length_plot <- ggplot(book_dependency_stats,
  # Sort by mean dependency length
  aes(x = reorder(book_id, mean_dep_length),
      y = mean_dep_length)) +
  geom_col(fill = "#5D93E1") +
  # Add error bars showing variability
  geom_errorbar(
    aes(ymin = mean_dep_length - median_dep_length/4, 
        ymax = mean_dep_length + median_dep_length/4),
    width = 0.2, color = "#2C528C"
  ) +
  labs(
    #title = "Syntactic Complexity in Gaiman's Works",
    subtitle = "Mean dependency length (average distance between related words)",
    x = "Book",
    y = "Mean Dependency Length"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )

# 5. Find illustrative examples of short and long dependencies
# We'll find examples from the original parsed sentences
parsed_with_dep_length <- parsed_samples %>%
  # Calculate dependency length for each token
  mutate(
    token_id_num = as.numeric(token_id),
    head_id_num = as.numeric(head_token_id),
    dep_length = abs(token_id_num - head_id_num)
  )

# Aggregate to get max dependency length in each sentence
sentence_max_deps <- parsed_with_dep_length %>%
  group_by(doc_id, sentence_id) %>%
  summarize(
    max_dep = max(dep_length[upos != "PUNCT" & head_token_id != "0"], na.rm = TRUE),
    sentence_text = first(sentence),
    .groups = "drop"
  ) %>%
  # Remove potential NA/Inf values from empty sentences
  filter(is.finite(max_dep))

# Find better examples
short_example <- sentence_max_deps %>%
  filter(max_dep > 1 & max_dep <= 3) %>%  # Not too trivial, but still simple
  filter(nchar(sentence_text) > 10) %>%    # Ensure it's a real sentence
  arrange(max_dep) %>%
  slice(1)

long_example <- sentence_max_deps %>%
  filter(max_dep >= 7) %>%                # Clearly complex
  arrange(desc(max_dep)) %>%
  slice(1)

# Join with book information for examples
example_sentences <- bind_rows(
  short_example %>% mutate(type = "Short dependency"),
  long_example %>% mutate(type = "Long dependency")
) %>%
  left_join(sample_book_mapping, by = c("doc_id" = "sample_id"))

# Print results
print(dep_length_plot)

# Print example sentences with book sources
#cat("\nExample sentences illustrating dependency length:\n\n")
# not very helpful and takes too much space, not use

#for(i in 1:nrow(example_sentences)) {
  #cat(paste0(example_sentences$type[i], " (max = ", 
             #round(example_sentences$max_dep[i], 1), 
             #") from ", example_sentences$title[i], ":\n"))
  #cat(example_sentences$sentence_text[i], "\n\n")
#}
```

#### Readability Analysis

The Flesch-Kincaid Grade Level scores provide standardised readability measurements based on word and sentence characteristics. @fig-readability_score shows these scores across Gaiman's works calculated by `quanteda` [@benoit2018] .

```{r}
#| label: fig-readability_score
#| echo: false
#| fig-cap: Reading Difficulty of Gaiman's Works
# Focus solely on Flesch-Kincaid Grade Level without imposing age categories

# 1. Calculate readability scores for each book
book_texts <- gaiman_corpus$document_level %>%
  group_by(book_id) %>%
  summarise(text = paste(text, collapse = " ")) %>%
  ungroup()

# 2. Use quanteda for readability calculations
readability_scores <- textstat_readability(
  book_texts$text,
  measure = c("Flesch.Kincaid")  # Focus on just this measure
)
readability_scores$book_id <- book_texts$book_id

# 3. Join with book information
readability_data <- readability_scores %>%
  # Get book information (title only)
  left_join(
    gaiman_corpus$metadata$book_level %>% 
      select(book_id, title), 
    by = "book_id"
  )

# 4. Create a cleaner, data-driven visualization
readability_plot <- ggplot(readability_data,
  # Sort by Flesch-Kincaid score (reading difficulty)
  aes(x = reorder(book_id, Flesch.Kincaid),
      y = Flesch.Kincaid)) +
  # Use a gradient color reflecting complexity, but not imposing categories
  geom_col(aes(fill = Flesch.Kincaid)) +
  scale_fill_gradient(low = "#E3F2FD", high = "#1565C0") +
  # Add labeled reference lines for grade level interpretation
  geom_hline(yintercept = c(5, 8, 12), linetype = "dashed", color = "gray60") +
  annotate("text", x = 1, y = 5.2, label = "5th Grade", hjust = 0, size = 3, color = "gray40") +
  annotate("text", x = 1, y = 8.2, label = "8th Grade", hjust = 0, size = 3, color = "gray40") +
  annotate("text", x = 1, y = 12.2, label = "12th Grade", hjust = 0, size = 3, color = "gray40") +
  labs(
    #title = "Reading Difficulty of Gaiman's Works",
    subtitle = "Flesch-Kincaid Grade Level (higher = more difficult reading)",
    x = "Book",
    y = "Reading Grade Level",
    fill = "Grade Level"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )

# 5. Add book titles to the plot for clearer identification
readability_plot <- readability_plot +
  # Add book titles as text labels
  geom_text(
    aes(label = title, y = 0.5),
    angle = 90, hjust = 0, size = 2.5, color = "grey45"
  )

# Display the visualization
print(readability_plot)
```

The results demonstrate a unexpected patterns that challenge conventional assumptions about audience targeting. All books are below 8th grade which fits Gaiman's niche as a writer of popular fiction. *American Gods*, often categorised as an adult fiction, scores lower than *The Graveyard Book*, which is tagged as "Middle Grade" and "Young Adult" on goodreads. This suggests that readability scores alone cannot reliably predict audience targeting in Gaiman's work.

This finding is echoed by recent criticism of readability metrics. @tanprasert2021 argue that "Flesch-Kincaid Grade Level (FKGL) should not be used to evaluate text simplification systems," noting that the score can be easily manipulated with minor textual changes. FKGL formulas primarily rely on surface features such as word and sentence length, while fail to capture semantic complexity, narrative sophistication, thematic maturity, or cultural references that can better differentiate works for different audiences.

#### Profanity Analysis

A dictionary-based approach was used to detect and categorise profanity across words. Three categories were established:

-   **mild profanity**: for example, "damn"

-   **moderate profanity**: for example, "ass, shit"

-   **strong profanity**: for example, "fuck, cunt"

The complete list of profanity can be found in Appendix. @fig-swear_analysis shows the frequency of profanity per 1,000 words for each book.

```{r}
#| label: swear_analysis_calculation
#| include: false

# 2. Define categorized profanity dictionaries
# Expanded lists with more comprehensive coverage
mild_profanity <- c(
  "damn", "darn", "hell", "crap", "suck", "stupid", "idiot", "dumb", 
  "moron", "fool", "jerk", "dork", "heck", "gosh", "jeez", "Christ", 
  "God", "Jesus", "bloody", "blast", "drat", "freaking", "frigging",
  "poop", "butt", "arse", "bollocks", "turd", "bum"
)

moderate_profanity <- c(
  "ass", "asshole", "bastard", "shit", "bullshit", "piss", 
  "screw", "screwed", "dick", "cock", "prick", "slut", "whore", 
  "wanker", "tosser", "twat", "crap", "balls", "nuts", "pissed", 
  "fart", "jackass", "douche", "douchebag", "tits", "boobs"
)

strong_profanity <- c(
  "fuck", "fucked", "fucker", "fucking", "motherfucker", "motherfucking",
  "cunt", "pussy", "cock", "bitch", "cum", "jizz", "nigger", "faggot", 
  "fag", "spic", "chink", "kike", "retard", "whore", "slut"
)

# Combine all for overall profanity detection
all_profanity <- unique(c(mild_profanity, moderate_profanity, strong_profanity))

# 3. Prepare sentence-level data
sentence_data <- gaiman_corpus$sentence_level %>%
  select(book_id, title, sentence_id, text)

# 4. Apply overall profanity detection 
profanity_results <- profanity_by(
  text.var = sentence_data$text,
  by = sentence_data$book_id,
  profanity_list = all_profanity
)

# 5. Create function to count profanity by category
count_profanity_by_category <- function(book_id, sentence_data) {
  # Extract text for this book
  book_text <- sentence_data %>%
    filter(book_id == !!book_id) %>%
    pull(text) %>%
    paste(collapse = " ") %>%
    tolower()
  
  # Count instances of each category
  mild_count <- sum(sapply(mild_profanity, function(term) {
    str_count(book_text, paste0("\\b", term, "\\b"))
  }))
  
  moderate_count <- sum(sapply(moderate_profanity, function(term) {
    str_count(book_text, paste0("\\b", term, "\\b"))
  }))
  
  strong_count <- sum(sapply(strong_profanity, function(term) {
    str_count(book_text, paste0("\\b", term, "\\b"))
  }))
  
  # Return counts by category
  return(c(
    mild = mild_count,
    moderate = moderate_count,
    strong = strong_count,
    total = mild_count + moderate_count + strong_count
  ))
}

# 6. Apply category counting to each book
book_ids <- unique(sentence_data$book_id)
category_counts <- lapply(book_ids, function(id) {
  counts <- count_profanity_by_category(id, sentence_data)
  data.frame(
    book_id = id,
    mild_count = counts["mild"],
    moderate_count = counts["moderate"],
    strong_count = counts["strong"],
    total_count = counts["total"]
  )
})

# Combine results
profanity_categories <- bind_rows(category_counts) %>%
  # Join with word counts from profanity_results
  left_join(
    profanity_results %>% select(book_id, word_count),
    by = "book_id"
  ) %>%
  # Calculate rates per 1000 words
  mutate(
    mild_per_1000 = (mild_count / word_count) * 1000,
    moderate_per_1000 = (moderate_count / word_count) * 1000,
    strong_per_1000 = (strong_count / word_count) * 1000,
    total_per_1000 = (total_count / word_count) * 1000
  ) %>%
  # Sort by total profanity rate
  arrange(desc(total_per_1000))

# 7. Create visualization of profanity categories
# Prepare data for stacked bar chart
profanity_long <- profanity_categories %>%
  select(book_id, ends_with("per_1000")) %>%
  # Keep only category columns
  select(-total_per_1000) %>%
  # Convert to long format
  pivot_longer(
    cols = ends_with("per_1000"),
    names_to = "category",
    values_to = "rate_per_1000"
  ) %>%
  # Clean up category names
  mutate(
    category = case_when(
      category == "mild_per_1000" ~ "Mild",
      category == "moderate_per_1000" ~ "Moderate", 
      category == "strong_per_1000" ~ "Strong"
    ),
    # Convert to factor with desired order
    category = factor(category, levels = c("Mild", "Moderate", "Strong"))
  )

# Create stacked bar chart
category_plot <- ggplot(profanity_long, 
                       aes(x = reorder(book_id, rate_per_1000, sum), 
                           y = rate_per_1000,
                           fill = category)) +
  geom_col() +
  scale_fill_manual(values = c(
    "Mild" = "#FFC107",
    "Moderate" = "#FF9800", 
    "Strong" = "#F44336"
  )) +
  labs(
    #title = "Profanity in Gaiman's Works by Intensity Level",
    subtitle = "Words per 1,000 categorized by severity",
    x = "Book",
    y = "Frequency (per 1,000 words)",
    fill = "Profanity Level"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )

# 8. Function to extract examples for each category
extract_category_examples <- function(book_id, category, word_list, n_examples = 2) {
  # Get sentences for this book
  book_sentences <- sentence_data %>%
    filter(book_id == !!book_id)
  
  # Find sentences containing words from the specified category
  matches <- list()
  for (term in word_list) {
    pattern <- paste0("\\b", term, "\\b")
    for (i in 1:nrow(book_sentences)) {
      text <- book_sentences$text[i]
      if (str_detect(tolower(text), pattern)) {
        # Create highlighted text with term in asterisks
        highlighted <- str_replace_all(
          text, 
          regex(pattern, ignore_case = TRUE),
          paste0("**", term, "**")
        )
        matches <- c(matches, list(list(
          sentence_id = book_sentences$sentence_id[i],
          text = highlighted,
          term = term
        )))
        # Stop if we have enough examples
        if (length(matches) >= n_examples) break
      }
    }
    # Stop if we have enough examples
    if (length(matches) >= n_examples) break
  }
  
  # Convert list to data frame
  if (length(matches) > 0) {
    result <- do.call(rbind, lapply(matches, function(x) {
      data.frame(
        sentence_id = x$sentence_id,
        text = x$text,
        term = x$term,
        stringsAsFactors = FALSE
      )
    }))
    return(head(result, n_examples))
  } else {
    return(data.frame(
      sentence_id = integer(0),
      text = character(0),
      term = character(0)
    ))
  }
}

# 9. Extract examples for top books
# Get top 3 books by total profanity
top_books <- profanity_categories %>%
  top_n(3, total_per_1000) %>%
  pull(book_id)

# Extract examples for each category from each top book
profanity_examples <- list()
for (book in top_books) {
  profanity_examples[[book]] <- list(
    mild = extract_category_examples(book, "Mild", mild_profanity),
    moderate = extract_category_examples(book, "Moderate", moderate_profanity),
    strong = extract_category_examples(book, "Strong", strong_profanity)
  )
}


```

```{r}
#| label: fig-swear_analysis
#| echo: false
#| fig-cap: Profanity in Gaiman's Works by Intensity Level

print(category_plot)


```

```{r}
#| label: swear_analysis_stats
#| include: false
# Print category statistics
cat("\nProfanity statistics by category and book:\n")
print(profanity_categories %>% 
      select(book_id, mild_per_1000, moderate_per_1000, strong_per_1000, total_per_1000))

```

```{r}
#| label: swear_analysis_examples
#| message: false
#| warning: false
#| include: false
#| paged-print: false
# 11. Print examples with category context
cat("\nExamples of profanity by category from top books:\n\n")
for (book in top_books) {
  cat("===== Examples from", book, "=====\n")
  
  # Print mild examples
  cat("\nMILD PROFANITY EXAMPLES:\n")
  examples <- profanity_examples[[book]]$mild
  if (nrow(examples) > 0) {
    for (i in 1:nrow(examples)) {
      cat(i, ": ", examples$text[i], "\n")
    }
  } else {
    cat("No mild profanity examples found.\n")
  }
  
  # Print moderate examples
  cat("\nMODERATE PROFANITY EXAMPLES:\n")
  examples <- profanity_examples[[book]]$moderate
  if (nrow(examples) > 0) {
    for (i in 1:nrow(examples)) {
      cat(i, ": ", examples$text[i], "\n")
    }
  } else {
    cat("No moderate profanity examples found.\n")
  }
  
  # Print strong examples
  cat("\nSTRONG PROFANITY EXAMPLES:\n")
  examples <- profanity_examples[[book]]$strong
  if (nrow(examples) > 0) {
    for (i in 1:nrow(examples)) {
      cat(i, ": ", examples$text[i], "\n")
    }
  } else {
    cat("No strong profanity examples found.\n")
  }
  
  cat("\n")
}
# No there is no space for these much examples
```

The data reveals notable variations in profanity usage across Gaiman corpus. *American Gods* shows the highest overall profanity rate (2.28 per 1,000 words), followed by *Smoke and Mirrors* (1.32 per 1,000 words), and *Neverwhere* (1.02). At the lower end, *Fortunately, the Milk* (0.13)*, Coraline* (0.19)*,* and *The Ocean at the End of the Lane* (0.24) contain minimal profanity. The figure also indicates the frequency and intensity of profanity are positively associated. The patterns of profanity distribution directly reflect the features that can influence demographic targeting such as thematic concerns, narrative tone, and publication context.

#### Sentiment Analysis

I calculated sentence-level sentiment variation by `sentimentr` [@tylerw2021]. @fig-senti_dist shows the sentiment distribution across books.

```{r}
#| label: fig-senti_dist
#| echo: false
#| message: false
#| warning: false
#| results: hide
#| fig-cap: "Sentiment Distribution in Gaiman's Works"

# First, let's set up a simple analysis pipeline

# 1. Calculate sentence-level sentiment
sentence_sentiment <- sentiment_by(
  text.var = sentence_data$text,
  by = list(
    book_id = sentence_data$book_id,
    sentence_id = sentence_data$sentence_id
  )
)

# 2. Classify each sentence as positive, negative, or neutral
sentence_sentiment <- sentence_sentiment %>%
  mutate(
    sentiment_class = case_when(
      ave_sentiment > 0.05 ~ "Positive",
      ave_sentiment < -0.05 ~ "Negative",
      TRUE ~ "Neutral"  # Use a small buffer around zero for neutral
    )
  )

# 3. Calculate the distribution for each book
book_sentiment_distribution <- sentence_sentiment %>%
  group_by(book_id) %>%
  summarize(
    total_sentences = n(),
    positive_count = sum(sentiment_class == "Positive"),
    negative_count = sum(sentiment_class == "Negative"),
    neutral_count = sum(sentiment_class == "Neutral"),
    positive_pct = (positive_count / total_sentences) * 100,
    negative_pct = (negative_count / total_sentences) * 100,
    neutral_pct = (neutral_count / total_sentences) * 100
  )

# 4. Verify our percentages add up to 100%
book_sentiment_distribution <- book_sentiment_distribution %>%
  mutate(total_pct = positive_pct + negative_pct + neutral_pct)

# Check results (should be close to 100% for each book)
print(book_sentiment_distribution[, c("book_id", "total_pct")])

# 5. Create data for visualization
sentiment_long <- book_sentiment_distribution %>%
  select(book_id, positive_pct, negative_pct, neutral_pct) %>%
  pivot_longer(
    cols = c("positive_pct", "negative_pct", "neutral_pct"),
    names_to = "sentiment",
    values_to = "percentage"
  ) %>%
  mutate(
    sentiment = case_when(
      sentiment == "positive_pct" ~ "Positive",
      sentiment == "negative_pct" ~ "Negative",
      sentiment == "neutral_pct" ~ "Neutral"
    )
  )

# 6. Create a simple stacked bar chart
sentiment_distribution_plot <- ggplot(sentiment_long, 
                                     aes(x = book_id, y = percentage, fill = sentiment)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c(
    "Positive" = "#4CAF50",
    "Negative" = "#F44336",
    "Neutral" = "#9E9E9E"
  )) +
  labs(
    #title = "Sentiment Distribution in Gaiman's Works",
    subtitle = "Percentage of positive, negative, and neutral sentences in each book",
    x = "Book",
    y = "Percentage of Sentences",
    fill = "Sentiment"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  ) +
  # Ensure y-axis goes from 0-100%
  scale_y_continuous(limits = c(0, 100))

# Display the plot
print(sentiment_distribution_plot)

# Optional: Generate a summary table for reporting
sentiment_summary <- book_sentiment_distribution %>%
  select(book_id, positive_pct, negative_pct, neutral_pct) %>%
  arrange(desc(positive_pct))

print(sentiment_summary)
```

The analysis demonstrates considerable consistency in sentiment distribution across Gaiman's works. Most books maintain a similar balance of emotional content (approximately 23 - 31% positive, 18 - 27% negative, and 43 - 52% neutral). *Odd and the Frost Giants* shows the highest proportion of negative sentences (33.45%) while *Stardust* shows the highest proportion of positive sentences (31%).

The emotional arcs — how sentiment progresses through the narrative – show some variation. @fig-senti_arc illustrates these emotional trajectories. *Smoke and Mirrors* (SM) shows the highest sentiment range (0.46 with starting sentiment at -0.03 and ends at 0.23). However, this pattern likely stems from its nature as collection of short stories rather than representing a stylistic choice related to audience targeting. In general the sentiment arcs of the ten books vary from -0.23 to 0.24. The overall stable sentiment distribution across works suggests that one the one hand, Gaiman maintains a consistent emotional palette regardless of intended readership; on the other hand, `sentimentr`, as a dictionary-based approach, may fail to catch some subtle sentiments such as irony and satire. Unlike profanity usage and syntactic complexity, sentiment patterns do not appear to correlate meaningfully with potential audience demographics.

In sentence-level analysis, profanity usage appears to be the most distinctive content-based marker, with clear patterns of variation across different works. They create groupings that partially but not perfectly align with traditional audience categorisations. The sentence-level features provides more nuanced insights than lexical analysis alone, suggesting that Gaiman's stylistic adaptations for different audiences operate more at the structural and content-appropriateness levels than through vocabulary simplification. The findings so far pave the way to examine discourse-level patterns to develop a more comprehensive understanding of potential audience-based stylistic variations.

```{r}
#| label: fig-senti_arc
#| echo: false
#| results: hide
#| fig-cap: "Emotional Arcs in Neil Gaiman's Works"
# Sentiment Flow Analysis
# This function visualizes how sentiment changes throughout each book's narrative progression

# 1. Calculate sentiment by chapter for each book
chapter_sentiment <- sentence_sentiment %>%
  # Join with document information to get chapter numbers
  left_join(
    gaiman_corpus$sentence_level %>% 
      select(book_id, sentence_id, chapter_num) %>% 
      distinct(),
    by = c("book_id", "sentence_id")
  ) %>%
  # Group by book and chapter to calculate average sentiment
  group_by(book_id, chapter_num) %>%
  summarize(
    chapter_sentiment = mean(ave_sentiment, na.rm = TRUE),
    sentences_count = n(),
    .groups = "drop"
  )

# 2. Calculate normalized position for each chapter within its book
normalized_sentiment_flow <- chapter_sentiment %>%
  # Group by book to calculate relative positions
  group_by(book_id) %>%
  mutate(
    # Get the max chapter number for each book
    max_chapter = max(chapter_num, na.rm = TRUE),
    # Calculate normalized position (0-100% of narrative)
    narrative_position = (chapter_num / max_chapter) * 100
  ) %>%
  # Get book titles for better labeling
  left_join(
    gaiman_corpus$document_level %>% 
      select(book_id, title) %>% 
      distinct(),
    by = "book_id"
  ) %>%
  # Make sure we don't have NA values
  filter(!is.na(chapter_sentiment))

# 3. Create a combined visualization showing all books' arcs
emotional_arcs_plot <- ggplot(normalized_sentiment_flow,
                             aes(x = narrative_position,
                                 y = chapter_sentiment,
                                 color = book_id)) +
  # Add smoothed trend lines to see the emotional arc
  geom_smooth(method = "loess", se = FALSE, span = 0.6, linewidth = 1) +
  # Add small points for actual chapter sentiments
  geom_point(alpha = 0.4, size = 1.5) +
  # Add horizontal reference line at neutral sentiment (0)
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.5) +
  # Add labels
  labs(
    #title = "Emotional Arcs in Neil Gaiman's Works",
    subtitle = "Sentiment progression throughout narrative (0% = beginning, 100% = end)",
    x = "Narrative Progression (%)",
    y = "Average Sentiment Score",
    color = "Book"
  ) +
  # Use a color palette that distinguishes between books
  scale_color_brewer(palette = "Paired") +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )

# 4. Create a facetted version showing individual book arcs
individual_arcs_plot <- ggplot(normalized_sentiment_flow,
                              aes(x = narrative_position,
                                  y = chapter_sentiment)) +
  # Add smoothed trend lines to see the emotional arc
  geom_smooth(method = "loess", se = FALSE, span = 0.6, aes(color = book_id), linewidth = 1) +
  # Add small points for actual chapter sentiments
  geom_point(alpha = 0.4, size = 1.5, color = "darkgray") +
  # Add horizontal reference line at neutral sentiment (0)
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.5) +
  # Facet by book title for individual arcs
  facet_wrap(~ title, scales = "free_y") +
  # Add labels
  labs(
    title = "Individual Emotional Arcs in Neil Gaiman's Works",
    subtitle = "Sentiment progression by narrative position for each book",
    x = "Narrative Progression (%)",
    y = "Average Sentiment Score"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.spacing = unit(1, "lines"),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# 5. Print both visualizations
print(emotional_arcs_plot)
#print(individual_arcs_plot) no need for reporting

# 6. Calculate key statistics about the arcs
arc_statistics <- normalized_sentiment_flow %>%
  group_by(book_id, title) %>%
  summarize(
    start_sentiment = first(chapter_sentiment),
    end_sentiment = last(chapter_sentiment),
    min_sentiment = min(chapter_sentiment),
    max_sentiment = max(chapter_sentiment),
    sentiment_range = max_sentiment - min_sentiment,
    overall_trend = end_sentiment - start_sentiment,
    .groups = "drop"
  ) %>%
  mutate(
    arc_type = case_when(
      overall_trend > 0.05 ~ "Rising (positive resolution)",
      overall_trend < -0.05 ~ "Falling (negative resolution)",
      TRUE ~ "Neutral/Circular"
    )
  ) %>%
  arrange(desc(overall_trend))

# 7. Print summary statistics
print(arc_statistics)
```

### Discourse level

After examining word- and sentence-level patterns, this section moves further by exploring discourse-level characteristics in Gaiman's work. It focuses on how textual elements function together to create meaning across larger units of text. In literature stylistics, discourse features often reveal how authors construct narrative voice, manage perspective, and guide reader engagement – all of which may differ based on intended readership.

#### Point of View analysis

Point of view (POV) represents a fundamental aspect of narrative discourse. The analysis tracked personal pronoun usage across the 10 books in the corpus to identify patterns in narrative perspective. @fig-pov_composition shows the composition of first (I/me), second (you), and third (she/he/they) pronouns as percentages of total words across the corpus.

Books like *Fortunately, the Milk* (FTM) and *The Ocean at the End of the Lane* (OCEAN) show higher rates of first-person narration compared to books like *American Gods* (AM), *Coraline* (CL), and *Neverwhere* (NW). This suggest that POV alone is not a feature differentiating the target audience group of a book.

Perspective stability analysis, on the other hand, reveals clearer patterns that may related to audience considerations. @fig-pov_stable displays the narrative perspective stability index. It measures how consistently each book maintains its dominant perspective without shifts. Books generally considered appropriate for younger readers (FTM and OFG) show higher stability indices (0.72 - 0.8), while books aimed at older audiences (AG, AB) demonstrate more perspective shifting (0.53 - 0.58). However, ST, a books that usually considered for young adults, has the highest shifting score.

```{r}
#| label: pov_shift_analysis
#| echo: false
#| warning: false
 # POV Stability Analysis (Window-based approach)
# ---------------------------------------

# First, check what columns are actually available in your token data
# str(gaiman_corpus$token_level)

# Adjust the analysis to work with your corpus structure
pov_stability <- gaiman_corpus$token_level %>%
  # Filter for pronoun tokens using token_lower instead of upos
  filter(token_lower %in% c(
    # First person
    "i", "me", "my", "mine", "myself", "we", "us", "our", "ours", "ourselves",
    # Second person
    "you", "your", "yours", "yourself", "yourselves",
    # Third person
    "he", "him", "his", "himself", "she", "her", "hers", "herself",
    "it", "its", "itself", "they", "them", "their", "theirs", "themselves"
  )) %>%
  # Add perspective classification
  mutate(
    perspective = case_when(
      token_lower %in% c("i", "me", "my", "mine", "myself", "we", "us", "our", "ours", "ourselves") ~ "first",
      token_lower %in% c("you", "your", "yours", "yourself", "yourselves") ~ "second",
      TRUE ~ "third"
    )
  ) %>%
  # Group by book and sentence
  group_by(book_id, sentence_id) %>%
  # Determine predominant perspective for each sentence
  summarize(
    perspective = case_when(
      sum(perspective == "first") > 0 ~ "first",
      sum(perspective == "second") > 0 ~ "second",
      sum(perspective == "third") > 0 ~ "third",
      TRUE ~ "none"  # Fallback (shouldn't happen given our filter)
    ),
    .groups = "drop"
  ) %>%
  # Group sentences into windows to avoid name-pronoun alternation issues
  group_by(book_id) %>%
  mutate(
    # Create perspective windows (groups of consecutive sentences)
    window_id = ceiling(row_number() / 5)
  ) %>%
  # Get dominant perspective for each window
  group_by(book_id, window_id) %>%
  summarize(
    dominant_perspective = names(which.max(table(perspective))),
    .groups = "drop"
  ) %>%
  # Track shifts in dominant perspective
  group_by(book_id) %>%
  mutate(
    prev_perspective = lag(dominant_perspective),
    is_shift = dominant_perspective != prev_perspective & !is.na(prev_perspective)
  ) %>%
  # Calculate stability index and perspective composition
  summarize(
    windows = n(),
    shifts = sum(is_shift, na.rm = TRUE),
    stability_index = 1 - (shifts / (windows - 1)),  # Higher = more stable
    pct_first = mean(dominant_perspective == "first", na.rm = TRUE) * 100,
    pct_second = mean(dominant_perspective == "second", na.rm = TRUE) * 100,
    pct_third = mean(dominant_perspective == "third", na.rm = TRUE) * 100,
    .groups = "drop"
  )

# Get book titles for better visualization
pov_stability <- pov_stability %>%
  left_join(
    gaiman_corpus$document_level %>% 
      select(book_id, title) %>% 
      distinct(),
    by = "book_id"
  )


```

```{r}
#| label: fig-pov_stable
#| echo: false
#| message: false
#| fig-cap: Narrative Perspective Stability in Gaiman's Works
# Create perspective stability visualization
stability_plot <- ggplot(pov_stability, 
                        aes(x = reorder(book_id, stability_index),
                            y = stability_index)) +
  geom_col(fill = "#6D9EC1") +
  geom_text(aes(label = sprintf("%.2f", stability_index)), 
            vjust = -0.5, size = 3) +
  labs(
    #title = "Narrative Perspective Stability in Gaiman's Works",
    subtitle = "Higher values indicate more consistent perspective throughout the work",
    x = "Book",
    y = "Perspective Stability Index"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# Display both visualizations
print(stability_plot)

```

```{r}
#| label: fig-pov_composition
#| echo: false
#| warning: false
#| fig-cap: Narrative Perspective Composition in Gaiman's Works
# Create perspective composition visualization
composition_data <- pov_stability %>%
  select(book_id, pct_first, pct_second, pct_third) %>%
  pivot_longer(
    cols = c(pct_first, pct_second, pct_third),
    names_to = "perspective",
    values_to = "percentage"
  ) %>%
  mutate(
    perspective = case_when(
      perspective == "pct_first" ~ "First Person",
      perspective == "pct_second" ~ "Second Person",
      perspective == "pct_third" ~ "Third Person"
    )
  )

composition_plot <- ggplot(composition_data,
                          aes(x = reorder(book_id, percentage, sum),
                              y = percentage,
                              fill = perspective)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("First Person" = "#E57373",
                               "Second Person" = "#FFB74D",
                               "Third Person" = "#81C784")) +
  labs(
    #title = "Narrative Perspective Composition in Gaiman's Works",
    subtitle = "Percentage of narrative told in each perspective",
    x = "Book",
    y = "Percentage",
    fill = "Perspective"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(composition_plot)
```

#### Modal verbs analysis

Based on the framework of modal grammar [@simpson2003], I categorised modal verbs into four functional groups:

1.  **World Building** (must, should, have to, always, never): These establish rules, obligations, and certainties within the narrative world.

2.  **Possibility** (may, might, could, can, maybe): These express uncertainty and potential options.

3.  **Hypothetical** (would, if, whether): These create conditional or speculative scenarios.

4.  **Intent** (will, shall, going to, want to): These signal future actions and determinations.

The complete list can be found in Appendix. The distribution of the four groups of modals across the books is shown in @fig-modal_verb.

The most striking difference appears in the "authority ratio" (@fig-authority) —the ratio of world-building and intent modals (groups 1 and 4, expressing certainty) to possibility and hypothetical modals (groups 2 and 3 which express uncertainty). *Fortunately, the Milk* (FTM) stands out with a distinctly high authority ratio (0.96), substantially higher than all other books which range between 0.42-0.53. This further challenges the attempt to categorise a work into a dichotomy of children/adults literature based on linguistic features alone. In fact, books like *Coraline* (0.45) and *The Graveyard Book* (0.45) have authority ratios comparable to or even lower than typically presumed adult-oriented works like *American Gods* (0.53) and *Smoke and Mirrors* (0.53).

```{r}
#| label: mocal_verb_cal
#| include: false
# Modal Verb Analysis by Narrative Function
# ---------------------------------------

# Prepare book texts - create if not already available
book_texts <- gaiman_corpus$document_level %>%
  group_by(book_id) %>%
  summarise(text = paste(text, collapse = " ")) %>%
  ungroup()

# Corrected function to categorize modal verbs by narrative function
analyze_modal_functions <- function(text) {
  # Define patterns for modal categories
  world_building_pattern <- "\\b(must|should|have to|has to|had to|always|never)\\b"
  possibility_pattern <- "\\b(may|might|could|can|maybe|perhaps)\\b"
  hypothetical_pattern <- "\\b(would|if|whether)\\b"
  intent_pattern <- "\\b(will|shall|going to|want to)\\b"
  
  # Count matches for each category
  world_count <- str_count(tolower(text), world_building_pattern)
  possibility_count <- str_count(tolower(text), possibility_pattern)
  hypothetical_count <- str_count(tolower(text), hypothetical_pattern)
  intent_count <- str_count(tolower(text), intent_pattern)
  
  # Return named vector of counts
  return(c(
    world_building = world_count,
    possibility = possibility_count,
    hypothetical = hypothetical_count,
    intent = intent_count
  ))
}

# Apply function to each book
modal_results <- book_texts %>%
  rowwise() %>%
  mutate(
    # Apply the function to get modal counts
    modal_counts = list(analyze_modal_functions(text)),
    # Extract individual counts
    world_building = modal_counts["world_building"],
    possibility = modal_counts["possibility"],
    hypothetical = modal_counts["hypothetical"],
    intent = modal_counts["intent"],
    # Get total word count for normalization
    total_words = str_count(text, "\\S+"),
    # Normalize per 1000 words
    world_building_per_k = (world_building / total_words) * 1000,
    possibility_per_k = (possibility / total_words) * 1000,
    hypothetical_per_k = (hypothetical / total_words) * 1000,
    intent_per_k = (intent / total_words) * 1000,
    # Calculate authority ratio (world_building + intent vs. possibility + hypothetical)
    authority_ratio = (world_building + intent) / (possibility + hypothetical)
  ) %>%
  # Select only needed columns
  select(book_id, 
         world_building, possibility, hypothetical, intent, 
         world_building_per_k, possibility_per_k, hypothetical_per_k, intent_per_k,
         authority_ratio)

```

```{r}
#| label: fig-modal_verb
#| echo: false
#| fig-cap: Modal Verb Usage by Narrative Function
# Get book titles for visualization
modal_results <- modal_results %>%
  left_join(
    gaiman_corpus$document_level %>% 
      select(book_id, title) %>% 
      distinct(),
    by = "book_id"
  )

# Prepare data for visualization
modal_long <- modal_results %>%
  select(book_id, title, ends_with("_per_k")) %>%
  pivot_longer(
    cols = ends_with("_per_k"),
    names_to = "modal_type",
    values_to = "frequency_per_k"
  ) %>%
  # Clean up category names
  mutate(
    modal_type = case_when(
      modal_type == "world_building_per_k" ~ "World Building",
      modal_type == "possibility_per_k" ~ "Possibility",
      modal_type == "hypothetical_per_k" ~ "Hypothetical",
      modal_type == "intent_per_k" ~ "Intent"
    )
  )

# Create improved modal verb usage visualization
modal_plot <- ggplot(modal_long,
                    aes(x = reorder(book_id, frequency_per_k, sum),
                        y = frequency_per_k,
                        fill = modal_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("World Building" = "#5C6BC0",
                               "Possibility" = "#26A69A",
                               "Hypothetical" = "#AB47BC",
                               "Intent" = "#EF5350")) +
  labs(
    title = "Modal Verb Usage by Narrative Function",
    subtitle = "Frequency per 1,000 words across Gaiman's works",
    x = "Book",
    y = "Frequency (per 1,000 words)",
    fill = "Modal Function"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(modal_plot)


```

```{r}
#| label: fig-authority
#| echo: false
#| fig-cap: Narrative Voice Authority in Gaiman's Works
# Authority ratio visualization
authority_plot <- ggplot(modal_results,
                        aes(x = reorder(book_id, authority_ratio),
                            y = authority_ratio)) +
  geom_col(fill = "#FF7043") +
  geom_text(aes(label = sprintf("%.2f", authority_ratio)), 
            vjust = -0.5, size = 3) +
  labs(
    #title = "Narrative Voice Authority in Gaiman's Works",
    subtitle = "Ratio of authoritative modals (world-building + intent) to uncertain modals (possibility + hypothetical)",
    x = "Book",
    y = "Authority Ratio"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Display visualizations

print(authority_plot)
```

This finding suggests that modal usage in Gaiman's work does not neatly align with presumed audience demographics. The exceptional authority ratio in *Fortunately, the Milk* may reflect its unique status as a humorous children's picture book rather than representing a general pattern for all children's literature. The similarity in authority ratios across most other books indicates that Gaiman maintains a relatively consistent approach to modal expression regardless of intended audience.

### Integrated Analysis

The previous sections examined Gaiman's stylistic variations at multiple linguistic levels. While individual analyses revealed noteworthy patterns, they also highlighted the limitations of single-dimension approaches to categorising works by intended audience. This section integrates findings across lexical, sentence, and discourse levels to develop a more comprehensive understanding of stylistic variation in Gaiman's works.

```{r}
#| label: fig-pca_feature_selection
#| echo: false
#| message: false
#| warning: false
#| results: hide 
#| fig-cap: "Linguistic Spectrum of Gaiman's Works"

# PCA Analysis of Linguistic Features in Neil Gaiman's Works
# Purpose: Data-driven exploration of stylistic patterns across books


# Step 1: Compile the features matrix
# -----------------------------------------------------------------
# Extract key metrics from previous analyses into a single dataframe
gaiman_features <- data.frame(
  book_id = gaiman_corpus$metadata$book_level$book_id,
  
  # Lexical features
  lexical_diversity = gaiman_corpus$metadata$book_level$lexical_diversity,
  avg_word_length = word_lengh_stats$avg_length,
  short_word_pct = word_lengh_stats$short_word_pct,
  long_word_pct = word_lengh_stats$long_word_pct,
  
  # Sentence features
  avg_sentence_length = sentence_length_stats$avg_length,
  short_sent_pct = sentence_length_stats$short_sent_pct,
  long_sent_pct = sentence_length_stats$long_sent_pct,
  mean_dep_length = book_dependency_stats$mean_dep_length,
  readability_score = readability_data$Flesch.Kincaid,
  
  # Sentiment and profanity
  positive_sentiment = book_sentiment_distribution$positive_pct,
  negative_sentiment = book_sentiment_distribution$negative_pct,
  total_profanity = profanity_categories$total_per_1000,
  strong_profanity = profanity_categories$strong_per_1000,
  
  # Discourse features
  first_person_pct = pov_stability$pct_first,
  third_person_pct = pov_stability$pct_third,
  pov_stability = pov_stability$stability_index,
  authority_ratio = modal_results$authority_ratio
)

# Step 2: Set book_id as rownames and remove from data matrix
row.names(gaiman_features) <- gaiman_features$book_id
gaiman_features <- gaiman_features %>% select(-book_id)

# Step 3: Run the PCA
# -----------------------------------------------------------------
# scale=TRUE ensures all features are standardized before PCA
gaiman_pca <- prcomp(gaiman_features, scale = TRUE)

# Step 4: Examine basic PCA results
# -----------------------------------------------------------------
# Variance explained by each principal component
#summary(gaiman_pca) 
#No raw data shown for report

# Extract the feature loadings on principal components
loadings <- gaiman_pca$rotation
# print(loadings[, 1:2])  # Loadings on first two components -- no show for final report 

# Step 5: Create a clear, informative visualization
book_scores <- as.data.frame(gaiman_pca$x)
book_scores$book_id <- rownames(book_scores)

# Join with book titles for clearer labeling
book_scores <- book_scores %>%
  left_join(
    gaiman_corpus$document_level %>%
      select(book_id, title) %>%
      distinct(),
    by = "book_id"
  )

# Extract feature loadings for arrows
loadings_df <- as.data.frame(gaiman_pca$rotation[, 1:2])
loadings_df$feature <- rownames(loadings_df)

# Create the enhanced biplot
pca_spectrum <- ggplot() +
  # Add feature vectors as arrows
  geom_segment(data = loadings_df,
               aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "darkblue", alpha = 0.7) +
  # Add feature labels
  geom_text_repel(data = loadings_df,
                  aes(x = PC1*5.5, y = PC2*5.5, label = feature),
                  size = 3, color = "darkblue") +
  # Add book points
  geom_point(data = book_scores, 
             aes(x = PC1, y = PC2),
             size = 3, color = "darkred") +
  # Add book labels
  geom_text_repel(data = book_scores,
                  aes(x = PC1, y = PC2, label = book_id),
                  size = 4, 
                  box.padding = 0.5,
                  point.padding = 0.3,
                  force = 10) +
  # Add interpretive annotations
  annotate("text", x = max(book_scores$PC1)*0.85, y = 0, 
           label = "More complex, advanced", 
           color = "darkblue", fontface = "bold", hjust = 1) +
  annotate("text", x = min(book_scores$PC1)*0.85, y = 0, 
           label = "Simpler, more accessible", 
           color = "darkblue", fontface = "bold", hjust = 0) +
  # Add titles and axis labels with percentage of variance explained
  labs(
    #title = "Linguistic Spectrum of Gaiman's Works",
    subtitle = "Books positioned by linguistic similarities",
    x = paste0("Linguistic Complexity (PC1: ", round(gaiman_pca$sdev[1]^2/sum(gaiman_pca$sdev^2)*100, 1), "%)"),
    y = paste0("Narrative Perspective (PC2: ", round(gaiman_pca$sdev[2]^2/sum(gaiman_pca$sdev^2)*100, 1), "%)")
  ) +
  # Visual enhancements
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "italic"),
    panel.grid.minor = element_blank()
  ) +
  # Add reference lines at origin
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7)

# Display the vis
print(pca_spectrum)


# Step 6: Identify key differentiating features
# -----------------------------------------------------------------
# Calculate feature contribution to first two principal components
feature_contrib <- data.frame(
  feature = colnames(gaiman_features),
  PC1_contribution = abs(loadings[, 1]),
  PC2_contribution = abs(loadings[, 2])
) %>%
  arrange(desc(PC1_contribution))

#print(feature_contrib)
# cant make it in the report maybe appendix
```

To synthesise the multidimensional linguistic data, this study employed Principal Component Analysis (PCA), a statistical technique that identifies patterns across multiple variables. PCA transforms potentially correlated variables into a smaller set of uncorrelated variables called principal components. Each component captures a portion of the total variation in the data, with the first component capturing the largest amount of variation, the second component capturing the second largest, and so on.

For this analysis, 12 key metrics from previous analyses were included:

-   **Lexical features**: lexical diversity, average word length, percentages of short and long words

-   **Sentence features**: average sentence length, percentages of short and long sentences, mean dependency length, readability scores

-   **Content features**: sentiment percentages (positive/negative), profanity frequencies

-   **Discourse features**: POV distributions, perspective stability, authority ratio

The PCA results reveal two distinct dimensions of stylistic variation in Gaiman's works. The first principal component, accounting for 42% of the total variance, corresponds primarily to linguistic complexity. This dimension shows substantial positive correlations with features that indicate more sophisticated language use: longer sentences (loading of 0.369), greater average word length (0.366), higher proportions of long words (0.307), and increased profanity usage (total: 0.351, strong: 0.295). Conversely, it correlates negatively with markers of simpler language, such as higher percentages of short words (-0.366) and shorter sentences (-0.323). These relationships suggest PC1 represents a spectrum from linguistically accessible text (negative values) to more complex expression (positive values).

The second principal component, explaining 28% of variance, captures variations in narrative perspective and approach. This dimension correlates positively with third-person narration (0.385), negative sentiment expression (0.339), lexical diversity (0.274), and syntactic complexity as measured by dependency length (0.250). In contrast, it shows negative associations with perspective stability (-0.405), first-person narration (-0.398), and higher authority ratios (-0.393). This pattern suggests PC2 distinguishes between two narrative approaches: at one end, stable first-person narration with higher narrative authority, and at the other, more varied third-person perspectives with greater lexical range and syntactic complexity.

Together, these two components provide a nuanced framework for understanding stylistic variation in Gaiman's corpus. Rather than revealing a single dimension of simplification for younger audiences, they suggest multiple independent aspects of stylistic adaptation. When plotted in this two-dimensional space (@fig-pca_feature_selection), Gaiman's works reveal patterns that both confirm and challenge conventional audience categorisations.

Along PC1 (Linguistic Complexity), *Fortunately, the Milk* appears to the right, indicating slightly greater complexity, while books like *Stardust*, *The Graveyard Book*, and *Ocean at the End of the Lane* cluster toward the left (simpler language). This contradicts expectations that children's books would consistently show simpler linguistic features. *American Gods* and *Anansi Boys* do show higher complexity as expected for adult literature, but *Smoke and Mirrors* appears with simpler linguistic features despite targeting adult readers.

PC2 (Narrative Perspective) shows additional differentiation, with *Fortunately, the Milk* at the extreme bottom (stable first-person narration with high authority), while *Anansi Boys* and *Stardust* occupy the upper regions (more third-person narration with higher negative sentiment). The *Ocean* *at the End of the Lane* and *Smoke and Mirrors* both exhibit lower PC2 values, suggesting similar narrative perspectives despite different target audiences.

These positions demonstrate that Gaiman's stylistic adaptations don't follow a simple children-adult dichotomy. Instead, they reflect specific creative choices tailored to each work's narrative requirements. For instance, *Fortunately, the Milk*'s position suggests Gaiman adopts complex linguistic features within a highly stable narrative framework for this children's book, perhaps reflecting its fantastical multiverse-travel plot. Meanwhile, *Ocean at the End of the Lane* uses simpler language to effectively convey its childlike perspective on adult themes.

This multidimensional analysis resolves apparent contradictions in individual metrics and reveals Gaiman as an author who adjusts specific stylistic dimensions based on narrative needs rather than making uniform simplifications for younger audiences.

# Discussion

# Conclusion

# Reference {#sec-reference}

::: {#refs}
:::

# Appendix {#sec-appendix-.appendix}

## Profanity Word List

1.  mild_profanity: "damn", "darn", "hell", "crap", "suck", "stupid", "idiot", "dumb", "moron", "fool", "jerk", "dork", "heck", "gosh", "jeez", "Christ", "God", "Jesus", "bloody", "blast", "drat", "freaking", "frigging", "poop", "butt", "arse", "bollocks", "turd", "bum"
2.  moderate_profanity: "ass", "asshole", "bastard", "shit", "bullshit", "piss", "screw", "screwed", "dick", "cock", "prick", "slut", "whore", "wanker", "tosser", "twat", "crap", "balls", "nuts", "pissed", "fart", "jackass", "douche", "douchebag", "tits", "boobs"
3.  strong_profanity: "fuck", "fucked", "fucker", "fucking", "motherfucker", "motherfucking", "cunt", "pussy", "cock", "bitch", "cum", "jizz", "nigger", "faggot", "fag", "spic", "chink", "kike", "retard", "whore", "slut"

## Modal Verbs 

-   world_building_pattern: must, should, have to, has to, had to, always, never

-   possibility_pattern: may, might, could, can, maybe, perhaps

-   hypothetical_pattern: would, if, whether

-   intent_pattern : will, shall, going to, want to"
