---
title: "Gaiman_Corpora_Building"
author: "Hongxu Zhou"
date: "2025-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
```

## Goal

The goal is to organise the single-book corpora into a comprehensive corpus. The notebook contains two parts: 1. corpus implementation, and 2. three-level structure (document -\> sentence -\> lexicon) design.

## Corpus Implementation

### Check data structure of each corpus

The metadata of each corpus may be of minor differences. I will check before merging them into one.

```{r}
# Load necessary libraries
library(dplyr)
library(purrr)
library(readr)
library(here)

# 1. Define the list of corpus files
corpus_files <- list.files(path = here("corpora"), 
                          pattern = "_corpus\\.rds$", 
                          full.names = TRUE)

# 2. Function to import and standardize each corpus
process_corpus <- function(file_path) {
  # Extract book_id from file name
  book_id <- gsub("_corpus\\.rds$", "", basename(file_path))
  
  # Import corpus
  corpus <- readRDS(file_path)
  
  # Handle different text column names
  # First check if we have "text" or "full_text" column
  if ("full_text" %in% names(corpus) && !("text" %in% names(corpus))) {
    # Rename full_text to text for consistency
    corpus <- corpus %>% rename(text = full_text)
  }
  
  # Select and standardize required fields
  standardized_corpus <- corpus %>%
    # Use any_of to avoid errors if a column is missing
    select(any_of(c("book_id", "title", "chapter_num", "nword", "nchar", "text"))) %>%
    # Ensure all required columns exist (create with NA if missing)
    mutate(
      book_id = if("book_id" %in% names(.)) book_id else book_id,
      title = if("title" %in% names(.)) title else NA_character_,
      chapter_num = if("chapter_num" %in% names(.)) as.integer(chapter_num) else NA_integer_,
      nword = if("nword" %in% names(.)) as.integer(nword) else NA_integer_,
      nchar = if("nchar" %in% names(.)) as.integer(nchar) else NA_integer_,
      text = as.character(text)
    )
  
  # Print info about the processed corpus for monitoring
  cat("Processed", book_id, "- Rows:", nrow(standardized_corpus), "\n")
  
  return(standardized_corpus)
}

# 3. Import and process all corpus files
all_corpora <- corpus_files %>%
  map_dfr(process_corpus)

# 4. Add a unique identifier for each document
all_corpora <- all_corpora %>%
  mutate(doc_id = paste0(book_id, "_ch", chapter_num)) %>%
  select(doc_id, book_id, title, chapter_num, nword, nchar, text)

# 5. Save the combined corpus
saveRDS(all_corpora, here("corpora", "gaiman_combined_corpus.rds"))

# Optional: Save as TSV for backup
write_tsv(all_corpora, here("corpora_backup", "gaiman_combined_corpus.tsv"))
```

Maintain data consistency by simplifying the title of Anasi Boys

```{r}



# Fix the inconsistent title format
cleaned_corpus <- combined_corpus %>%
  mutate(title = case_when(
    # Replace the specific problematic title
    title == "Gaiman, Neil - Anansi Boys" ~ "Anansi Boys",
    # Keep all other titles as they are
    TRUE ~ title
  ))

# Save the updated corpus
saveRDS(cleaned_corpus, here("corpora", "gaiman_combined_corpus.rds"))

# Update the backup TSV as well
write_tsv(cleaned_corpus, here("corpora_backup", "gaiman_combined_corpus.tsv"))
```

Another issue, the chapter 14 in Anasi Boys is wrong.

```{r}


# Load the combined corpus
combined_corpus <- readRDS(here("corpora", "gaiman_combined_corpus.rds"))

# Remove the duplicate AB_ch14 rows with very low word counts
# Rows 15, 17, and 18 have word counts of 21, 23, and 23 respectively
cleaned_corpus <- combined_corpus %>%
  # This filters out rows where book_id is "AB", chapter_num is 14, 
  # and the word count is very small (less than 100)
  filter(!(book_id == "AB" & chapter_num == 14 & nword < 100))

# Check the result - there should now be only one AB_ch14 entry
cleaned_corpus %>%
  filter(book_id == "AB" & chapter_num == 14) %>%
  select(doc_id, book_id, chapter_num, nword)

# Save the updated corpus
saveRDS(cleaned_corpus, here("corpora", "gaiman_combined_corpus.rds"))

# Update the backup TSV as well
write_tsv(cleaned_corpus, here("corpora_backup", "gaiman_combined_corpus.tsv"))
```
